---
title: "Interaction"
author: "Mark Lai"
date: "April 2, 2022"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", collapse = TRUE,
                      fig.width = 6, fig.asp = 0.618,
                      out.width = "70%", fig.align = "center")
comma <- function(x, digits. = 2L) {
    format(x, digits = digits., big.mark = ",")
}
```

```{r load-pkg, message = FALSE}
library(tidyverse)
library(here)
library(brms)  # simplify fitting Stan GLM models
library(posterior)  # for summarizing draws
library(modelsummary)  # table for brms
theme_set(theme_classic() +
    theme(panel.grid.major.y = element_line(color = "grey92")))
```

```{r waffle_divorce}
waffle_divorce <- read_delim(  # read delimited files
    "https://raw.githubusercontent.com/rmcelreath/rethinking/master/data/WaffleDivorce.csv",
    delim = ";"
)
# Rescale Marriage and Divorce by dividing by 10
waffle_divorce$Marriage <- waffle_divorce$Marriage / 10
waffle_divorce$Divorce <- waffle_divorce$Divorce / 10
waffle_divorce$MedianAgeMarriage <- waffle_divorce$MedianAgeMarriage / 10
# Recode `South` to a factor variable
waffle_divorce$South <- factor(waffle_divorce$South,
    levels = c(0, 1),
    labels = c("non-south", "south")
)
# See data description at https://rdrr.io/github/rmcelreath/rethinking/man/WaffleDivorce.html
```

# Different Slopes Across Two Groups

## Stratified Analysis

Let's consider whether the association between `MedianAgeMarriage` and `Divorce` differs between Southern and non-Southern states. Because (and **only because**) the groups are **independent**, we can fit a linear regression for each subset of states.

```{r smooth-south}
ggplot(waffle_divorce,
       aes(x = MedianAgeMarriage, y = Divorce, col = South)) +
    geom_point() +
    geom_smooth() +
    labs(x = "Median age marriage (10 years)",
         y = "Divorce rate (per 10 adults)") +
    ggrepel::geom_text_repel(aes(label = Loc))
```

```{r m_nonsouth, results = "hide"}
m_nonsouth <-
    brm(Divorce ~ MedianAgeMarriage,
        data = filter(waffle_divorce, South == "non-south"),
        prior = prior(normal(0, 2), class = "b") +
            prior(normal(0, 10), class = "Intercept") +
            prior(student_t(4, 0, 3), class = "sigma"),
        seed = 941,
        iter = 4000
    )
m_south <-
    brm(Divorce ~ MedianAgeMarriage,
        data = filter(waffle_divorce, South == "south"),
        prior = prior(normal(0, 2), class = "b") +
            prior(normal(0, 10), class = "Intercept") +
            prior(student_t(4, 0, 3), class = "sigma"),
        seed = 2157,  # use a different seed
        iter = 4000
    )
```

```{r tab-south}
msummary(list(South = m_south, `Non-South` = m_nonsouth),
         estimate = "{estimate} [{conf.low}, {conf.high}]",
         statistic = NULL, fmt = 2,
         gof_omit = "^(?!Num)"  # only include number of observations
)
```

We can now ask two questions:

- Is the intercept different across southern and non-southern states?
- Is the slope different across southern and non-southern states?

The correct way to answer the above questions is to obtain the posterior distribution of the **difference** in the coefficients. Repeat: obtain the posterior distribution of the **difference**. The incorrect way is to compare whether the CIs overlap.

Here are the posteriors of the differences:

```{r diff-south}
# Extract draws
draws_south <- as_draws_matrix(m_south,
    variable = c("b_Intercept", "b_MedianAgeMarriage")
)
draws_nonsouth <- as_draws_matrix(m_nonsouth,
    variable = c("b_Intercept", "b_MedianAgeMarriage")
)
# Difference in coefficients
draws_diff <- draws_south - draws_nonsouth
# Rename the columns
colnames(draws_diff) <- paste0("d", colnames(draws_diff))
# Summarize
summarize_draws(draws_diff)
```

As you can see, the southern states have a larger intercept and a lower slope.

```{r cond-south}
p1 <- plot(
    conditional_effects(m_nonsouth),
    points = TRUE, plot = FALSE
)[[1]] + ggtitle("South") + lims(x = c(2.3, 3), y = c(0.6, 1.4))
p2 <- plot(
    conditional_effects(m_south),
    points = TRUE, plot = FALSE
)[[1]] + ggtitle("Non-South") + lims(x = c(2.3, 3), y = c(0.6, 1.4))
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Interaction Model

An alternative is to include an interaction term

$$
  \begin{aligned}
    D_i & \sim N(\mu_i, \sigma)  \\
    \mu_i & = \beta_0 + \beta_1 S_i + \beta_2 A_i + \beta_3 S_i \times A_i \\
    \beta_0 & \sim N(0, 10) \\
    \beta_1 & \sim N(0, 10) \\
    \beta_2 & \sim N(0, 1) \\
    \beta_3 & \sim N(0, 2) \\
    \sigma & \sim t^+_4(0, 3)
  \end{aligned}
$$

- $\beta_1$: Difference in intercept between southern and non-southern states.
- $\beta_3$: Difference in the coefficient for A &rarr; D between southern and non-southern states

In the model, the variable S, southern state, is a dummy variable with 0 = non-southern and 1 = southern. Therefore, 

> For non-southern states, $\mu = (\beta_0) + (\beta_2) A$;
> for southern states, $\mu = (\beta_0 + \beta_1) + (\beta_2 + \beta_3) A$

```{r m1, results = "hide"}
m1 <- brm(
    Divorce ~ South * MedianAgeMarriage,
    data = waffle_divorce,
    prior = prior(normal(0, 2), class = "b") +
        prior(normal(0, 10), class = "b", coef = "Southsouth") +
        prior(normal(0, 10), class = "Intercept") +
        prior(student_t(4, 0, 3), class = "sigma"),
    seed = 941,
    iter = 4000
)
```

The formula `Divorce ~ South * MedianAgeMarriage` is the same as

```{r form1, eval = FALSE}
Divorce ~ South + MedianAgeMarriage + South:MedianAgeMarriage
```

where `:` is the symbol in R for a product term.

```{r print-m1}
m1
```

### Posterior predictive checks

```{r ppc-m1}
# Check density (normality)
pp_check(m1, type = "dens_overlay_grouped", group = "South")
# Check prediction (a few outliers)
pp_check(m1,
    type = "ribbon_grouped", x = "MedianAgeMarriage",
    group = "South"
)
# Check errors (no clear pattern)
pp_check(m1,
    type = "error_scatter_avg_vs_x", x = "MedianAgeMarriage"
)
```

### Conditional effects/simple slopes

Slope of `MedianAgeMarriage` when South = 0: $\beta_1$

Slope of `MedianAgeMarriage` when South = 1: $\beta_1 + \beta_3$

```{r summ-simple-slopes}
as_draws(m1) %>%
    mutate_variables(
        b_nonsouth = b_MedianAgeMarriage,
        b_south = b_MedianAgeMarriage + `b_Southsouth:MedianAgeMarriage`
    ) %>%
    posterior::subset_draws(
        variable = c("b_nonsouth", "b_south")
    ) %>%
    summarize_draws()
```

```{r cond-eff-m1}
plot(
    conditional_effects(m1,
        effects = "MedianAgeMarriage",
        conditions = data.frame(South = c("south", "non-south"),
                                cond__ = c("South", "Non-South"))
    ),
    points = TRUE
)
```

# Interaction of Continuous Predictors

```{r 3dscatter}
plotly::plot_ly(waffle_divorce,
                x = ~Marriage,
                y = ~MedianAgeMarriage,
                z = ~Divorce)
```

$$
  \begin{aligned}
    D_i & \sim N(\mu_i, \sigma)  \\
    \mu_i & = \beta_0 + \beta_1 M_i + \beta_2 A_i + \beta_3 M_i \times A_i \\
  \end{aligned}
$$

```{r m2, results = "hide"}
# Use default priors (just for convenience here)
m2 <- brm(Divorce ~ Marriage * MedianAgeMarriage,
    data = waffle_divorce,
    seed = 941,
    iter = 4000
)
```

## Centering

In the previous model, $\beta_1$ is the slope of M &rarr; D when A is 0 (i.e., median marriage age = 0), and $\beta_2$ is the slope of A &rarr; D when M is 0 (i.e., marriage rate is 0). These two are not very meaningful. Therefore, it is common to make the zero values more meaningful by doing *centering*.

Here, I use M - 2 as the predictor, so the zero point means a marriage rate of 2 per 10 adults; I use A - 2.5 as the other predictor, so the zero point means a median marriage rate of 25 years old.

$$\mu_i = \beta_0 + \beta_1 (M_i - 2) + \beta_2 (A_i - 2.5) + \beta_3 (M_i - 2) \times (A_i - 2.5)$$

```{r m2c, results = "hide"}
# Use default priors (just for convenience here)
m2c <- brm(Divorce ~ I(Marriage - 2) * I(MedianAgeMarriage - 2.5),
    data = waffle_divorce,
    seed = 941,
    iter = 4000
)
```

```{r tab-m2}
msummary(list(`No centering` = m2, `centered` = m2c),
         estimate = "{estimate} [{conf.low}, {conf.high}]",
         statistic = NULL, fmt = 2)
```

As shown in the table above, while the two models are equivalent in fit and give the same posterior distribution for $\beta_3$, they differ in $\beta_0$, $\beta_1$, and $\beta_2$.

```{r cond-eff-m2}
plot(
    conditional_effects(m2c,
        effects = "Marriage:MedianAgeMarriage",
        int_conditions = list(MedianAgeMarriage = c(2.3, 2.5, 2.7)),
    ),
    points = TRUE
)
```

# Multilevel Model

When data are naturally clustered in three or more segments or clusters, we can model interactions with a technique we have learned---hierarchical model with partial pooling. The difference is that we can have multiple parameters in each cluster. For example, consider the UC Berkeley admission data. 

```{r berkeley_admit}
berkeley_admit <- UCBAdmissions %>%
    as.data.frame() %>%
    group_by(Gender, Dept) %>%
    mutate(App = sum(Freq)) %>%
    filter(Admit == "Admitted") %>%
    ungroup() %>%
    select(Gender, Dept, Admit = Freq, App)
```

```{r plot-admit}
ggplot(berkeley_admit, aes(x = Gender)) +
    geom_pointrange(
        data = berkeley_admit %>%
            group_by(Gender) %>%
            summarise(
                padmit = sum(Admit) / sum(App),
                padmit_se = sqrt(padmit * (1 - padmit) / sum(App))
            ),
        aes(
            y = padmit,
            ymin = padmit - padmit_se, ymax = padmit + padmit_se
        )
    ) +
    labs(y = "Aggregated proportion admitted")
```

If we consider one department, we can model the number of admitted students for each gender as

$$
  \begin{aligned}
    z_i & \sim \text{Bin}(N, \mu_i)  \\
    \mathrm{logit}(\mu_i) & = \eta_i  \\
    \eta_i & = \beta_0 + \beta_1 \text{Gender}_i
  \end{aligned}
$$

So there are two coefficients, $\beta_0$ and $\beta_1$. We can then do the same for each of the six departments, and use partial pooling to pool the $\beta_0$'s into a common normal distribution, and the $\beta_1$'s into another common normal distribution. We can use $j$ = 1, 2, $\ldots$, $J$ to index department, and then we have the following multilevel model:

$$
  \begin{aligned}
    z_{ij} & \sim \text{Bin}(N_j, \mu_{ij})  \\
    \mathrm{logit}(\mu_{ij}) & = \eta_{ij}  \\
    \eta_{ij} & = \beta_{0j} + \beta_{1j} \text{Gender}_{ij}
  \end{aligned},
$$

and use a **multivariate normal** distribution to partially pool the $\beta_0$ and $\beta_1$ coefficients. The multivariate normal allows the $\beta_0$'s and $\beta_1$'s to be correlated:

$$\begin{bmatrix}
    \beta_{0j} \\
    \beta_{1j} \\
  \end{bmatrix} \sim N_2\left(
  \begin{bmatrix} 
    \gamma_0 \\
    \gamma_1 \\
  \end{bmatrix}, \mathbf{T}
  \right)$$

$N_2(\cdot)$ means a bivariate normal distribution, and $\mathbf{T}$ is a 2 $\times$ 2 covariance matrix for $\beta_0$ and $\beta_1$. To set priors for $\mathbf{T}$, we further decompose it into the standard deviations and the correlation matrix:

$$\mathbf{T} = 
  \begin{bmatrix}
    \tau_0 & 0 \\
    0 & \tau_1 \\
  \end{bmatrix}
  \begin{bmatrix}
    1 &  \\
    \rho_{10} & 1 \\
  \end{bmatrix}
  \begin{bmatrix}
    \tau_0 & 0 \\
    0 & \tau_1 \\
  \end{bmatrix}$$

We can use the same inverse-gamma or half-$t$ distributions for the $\tau$'s, as we've done in previous weeks. For $\rho$, we need to introduce a new distribution: the LKJ distribution.

## LKJ Prior

The LKJ Prior is a probability distribution for correlation matrices. A correlation matrix has 1 on all the diagonal elements. For example, a 2 $\times$ 2 correlation matrix is

$$\begin{bmatrix}
    1 & \\
    0.35 & 1
  \end{bmatrix}$$

where the correlation is 0.35. Therefore, with two variables, there is one correlation; with three or more variables, the number of correlations will be $q (q - 1) / 2$, where $q$ is the number of variables. 

For a correlation matrix of a given size, the LKJ prior has one shape parameter, $\eta$, where $\eta = 1$ corresponds to a uniform distribution of the correlations such that any correlations are equally likely, $\eta \geq 1$ favors a matrix closer to an identity matrix so that the correlations are closer to zero, and $\eta \leq 1$ favors a matrix with larger correlations. For a 2 $\times$ 2 matrix, the distribution of the correlation, $\rho$, with different $\eta$ values are shown in the graph below:

```{r plot-lkj}
dlkjcorr2 <- function(rho, eta = 1, log = FALSE) {
    # Function to compute the LKJ density given a correlation
    out <- (eta - 1) * log(1 - rho^2) -
        1 / 2 * log(pi) - lgamma(eta) + lgamma(eta + 1 / 2)
    if (!log) out <- exp(out)
    out
}
ggplot(tibble(rho = c(-1, 1)), aes(x = rho)) +
    stat_function(
        fun = dlkjcorr2, args = list(eta = 0.1),
        aes(col = "0.1"), n = 501
    ) +
    stat_function(
        fun = dlkjcorr2, args = list(eta = 0.5),
        aes(col = "0.5"), n = 501
    ) +
    stat_function(
        fun = dlkjcorr2, args = list(eta = 1),
        aes(col = "1"), n = 501
    ) +
    stat_function(
        fun = dlkjcorr2, args = list(eta = 2),
        aes(col = "2"), n = 501
    ) +
    stat_function(
        fun = dlkjcorr2, args = list(eta = 10),
        aes(col = "10"), n = 501
    ) +
    stat_function(
        fun = dlkjcorr2, args = list(eta = 100),
        aes(col = "100"), n = 501
    ) +
    labs(col = expression(eta), x = expression(rho), y = "Density")
```

As you can see, when $\eta$ increases, the correlation is more concentrated to zero. 

The default in `brms` is to use $\eta$ = 1, which is non-informative. If you have a weak but informative belief that the correlations shouldn't be very large, using $\eta$ = 2 is reasonable.

## Adding Cluster Means

In the multilevel modeling tradition, it is common also to include the **cluster means** of the within-cluster predictors. In this example, it means including the proportion of female applicants, `pFemale`. So the equation becomes

$$\eta_{ij} = \beta_{0j} + \beta_{1j} \text{Gender}_{ij} + \gamma_2 \text{pFemale}_j,$$

with one additional $\gamma_2$ coefficient (no $j$ subscript).

```{r pFemale}
# Obtain mean gender ratio at department level
berkeley_admit <- berkeley_admit %>%
    group_by(Dept) %>%
    mutate(pFemale = App[2] / sum(App)) %>%
    ungroup()
knitr::kable(berkeley_admit)
```

## Fitting the multilevel model in `brms`

For this example, I'll use these priors:

$$
  \begin{aligned}
    \gamma_0 & \sim t_4(0, 5) \\
    \gamma_1 & \sim t_4(0, 2.5) \\
    \gamma_2 & \sim t_4(0, 5) \\
    \tau_0 & \sim t^+_4(0, 3) \\
    \tau_1 & \sim t^+_4(0, 3) \\
    \rho & \sim \mathrm{LKJ}(2) \\
  \end{aligned},
$$

```{r m3, results = "hide"}
m3 <- brm(Admit | trials(App) ~ Gender + pFemale + (Gender | Dept),
    data = berkeley_admit,
    family = binomial("logit"),
    prior = prior(student_t(4, 0, 5), class = "Intercept") +
        prior(student_t(4, 0, 2.5), class = "b", coef = "GenderFemale") +
        prior(student_t(4, 0, 5), class = "sd") +
        prior(lkj(2), class = "cor"),
    seed = 1547,
    iter = 4000,
    # a larger adapt_delta usually needed for MLM
    control = list(adapt_delta = .99, max_treedepth = 12)
)
```

The estimated $\beta_0$ and $\beta_1$ for each department is

```{r coef-m3}
coef(m3)  # department-specific coefficients
```

And a posterior predictive check

```{r ppc-m3}
pp_check(m3, type = "intervals")
```

The plot below shows the predicted admission rate:

```{r fitted-m3}
berkeley_admit %>%
    bind_cols(fitted(m3)) %>%
    ggplot(aes(x = Dept, y = Admit / App,
               col = Gender)) +
    geom_errorbar(aes(ymin = `Q2.5` / App, ymax = `Q97.5` / App),
    position = position_dodge(0.3), width = 0.2) +
    geom_point(position = position_dodge(width = 0.3)) +
    labs(y = "Posterior predicted acceptance rate")
```

## Bonus: Growth Model

```{r sleepstudy}
data(sleepstudy, package = "lme4")
# Rescale reaction time
sleepstudy <- sleepstudy %>%
    mutate(Reaction100 = Reaction / 100)
```

- Data: Reaction times in a sleep deprivation study
- Predictor: Number of days of sleep deprivation
- Outcome: Daily average reaction time (ms)
- Cluster: 18 individuals (10 observations each)

$$
  \begin{aligned}
  \text{Repeated-measure level:}  \\
    \text{Reaction10}_{ij} & \sim \mathrm{lognormal}(\mu_{ij}, \sigma)  \\
    \mu_{ij} & = \beta_{0j} + \beta_{1j} \text{Days}_{ij}  \\
  \text{Person level:}  \\
    \begin{bmatrix}
      \beta_{0j} \\
      \beta_{1j} \\
    \end{bmatrix} & \sim N_2\left(
      \begin{bmatrix} 
        \gamma_0 \\
        \gamma_1 \\
      \end{bmatrix}, \mathbf{T}
      \right)  \\
    \mathbf{T} T & = \mathrm{diag}(\boldsymbol{\tau}) \boldsymbol{\Omega} \mathrm{diag}(\boldsymbol{\tau}) \\
  \text{Priors:}  \\
    \gamma_0 & \sim N(0, 2) \\
    \gamma_1 & \sim N(0, 1) \\
    \tau_0, \tau_1 & \sim t^+_4(0, 2.5) \\
    \boldsymbol{\Omega} & \sim \mathrm{LKJ}(2) \\
    \sigma & \sim t^+_4(0, 2.5)
  \end{aligned}
$$

```{r m4, results = "hide"}
m4 <- brm(
    Reaction100 ~ Days + (Days | Subject),
    data = sleepstudy,
    family = lognormal(),
    prior = c( # for intercept
        prior(normal(0, 2), class = "Intercept"),
        # for slope
        prior(std_normal(), class = "b"),
        # for tau0 and tau1
        prior(student_t(4, 0, 2.5), class = "sd"),
        # for correlation
        prior(lkj(2), class = "cor"),
        # for sigma
        prior(student_t(4, 0, 2.5), class = "sigma")
    ),
    control = list(adapt_delta = .95),
    seed = 2107,
    iter = 4000
)
```

```{r print-m4}
m4
```

Model estimate: the shaded band is the predicted mean trajectory

```{r pred-mean-traj, out.width = "100%", fig.width = 8.57}
sleepstudy %>%
    bind_cols(fitted(m4)) %>%
    ggplot(aes(x = Days, y = Reaction100)) +
    geom_ribbon(aes(y = Estimate, ymin = `Q2.5`,
                    ymax = `Q97.5`), alpha = 0.3) +
    geom_point() +
    facet_wrap(~ Subject)
```

<!-- The following chunk is only used on the website -->

```{r, echo = FALSE, results = 'asis', eval = file.exists("distill-hack.RMarkdown")}
knitrenv <- knitr::knit_global()
assign("knit_input", knitr::current_input(), knitrenv)
append_res <- knitr::knit_child("distill-hack.RMarkdown",
    envir = knitrenv, quiet = TRUE
)
cat(append_res)
```
