---
title: Modules
description: |
  Week learning objectives, slides, and task list
output:
  distill::distill_article:
    toc: false
---

```{r convert-xaringan, include = FALSE}
# Convert Rmd slides to HTML and PDF
for (f in dir("slides", pattern = "*.Rmd", full.names = TRUE)) {
  html_file <- xfun::with_ext(f, "html")
  pdf_file <- xfun::with_ext(f, "pdf")
  if (!file.exists(html_file) || file_test("-ot", html_file, f)) {
    xfun::Rscript_call(rmarkdown::render, args = list(input = f))
  }
  if (!file.exists(pdf_file) || file_test("-ot", pdf_file, f)) {
    pagedown::chrome_print(html_file)
  }
}
```

```{r, include = FALSE}
library(xaringanExtra, include.only = "embed_xaringan")
```

```{r panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

::: {.panelset}
::: {.panel}

# Week 1

## Week Learning Objectives

By the end of this module, you will be able to 

- Navigate the course website and Blackboard site
- Identify the Slack channels relevant for the course
- Describe the historical origin of Bayesian statistics
- Identify components in research papers involving Bayesian analyses
- Knit a simple R Markdown file

## Task Lists

1. Review the [syllabus](syllabus.html)
2. Review the resources (slides and [note][5])
3. Install/Update R and RStudio on your computer
4. Attend the Tuesday and Thursday class meetings
5. Complete the assigned readings
    * [Kruschke ch. 2][2]
    * Supplemental (i.e., optional) reading: [James ch. 1][3]
    * [Markdown Quick Reference][4]
6. Introduce yourself on the #introduction Slack channel (as part of HW 1)
7. Complete Homework 1 (see instruction on Blackboard)

[2]: https://drive.google.com/file/d/1tr7XpIVf04S44o5ZploxJcd37Fg1GygV/view
[3]: https://jim-stone.staff.shef.ac.uk/BookBayes2012/bookbayesch01.pdf
[4]: https://web.mit.edu/r/current/RStudio/resources/markdown_help.html
[5]: r_markdown.html

## Slides

[PDF version](slides/intro.pdf){target="_blank"}

```{r embed-1, echo = FALSE}
embed_xaringan("slides/intro.html", ratio = "4:3")
```

:::

::: {.panel}

# Week 2

## Week Learning Objectives

By the end of this module, you will be able to 

- Explain the **three axioms/properties** of probability
- Describe the **subjectivist** interpretation of probability, and contrast it with the **frequentist** interpretation
- Explain the difference between **probability mass** and **probability density**
- Compute probability density using **simulations**
- Compute **joint, marginal, and conditional probabilities** with two variables
- Write an R function and use loops for repeating computations

## Task Lists

1. Complete the assigned readings
    * Kruschke ch. 4
    * Wickham & Grolemund ch. [7][6], [19.1-19.5][7]
2. Review the resources (slides and notes on [probability][8] and [R basics][9])
3. Attend the Tuesday and Thursday class meetings
4. Complete Homework 2 (see instruction on Blackboard)

[6]: https://r4ds.had.co.nz/exploratory-data-analysis.html
[7]: https://r4ds.had.co.nz/functions.html
[8]: probability.html
[9]: r_basics.html

## Slides

[PDF version](slides/probability.pdf){target="_blank"}

```{r embed-2, echo = FALSE}
embed_xaringan("slides/probability.html", ratio = "4:3")
```

:::

::: {.panel}

# Week 3

## Week Learning Objectives

By the end of this module, you will be able to 

- Derive **Bayes' rule** from the definition of conditional probability
- Apply Bayes' rule to obtain **posterior** from prior and data
- Explain what **data-order invariance** and **exchangeability** are
- Use **grid approximation** to obtain the posterior for a **Bernoulli** model
- Describe the influence of **sample size** and **prior** on the posterior
- Use R to perform **prior predictive checks**

## Task Lists

1. Complete the assigned readings
    * [Gigerenzer (2004)][1]
    * Kruschke ch. 5
2. Review the resources (slides and [note][10])
3. Attend the Tuesday and Thursday class meetings
4. No homework for this week, but you may work on Q1 and Q2 for Homework 3 (see instruction on Blackboard)

[1]: https://pure.mpg.de/rest/items/item_2101336/component/file_2101335/content
[10]: bayes_rule.html

## Slides

<!--

[PDF version](slides/bayes_rule.pdf){target="_blank"}

```{r embed-3, echo = FALSE}
embed_xaringan("slides/bayes_rule.html", ratio = "4:3")
```

-->

:::

:::

## Last updated {.appendix}

```{r, echo = FALSE}
format(Sys.Date(), "%B %d, %Y")
```
