---
title: "Poisson Model"
author: "Mark Lai"
date: "January 30, 2022"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", collapse = TRUE,
                      fig.width = 6, fig.asp = 0.618,
                      out.width = "70%", fig.align = "center",
                      fig.show = "hold")
comma <- function(x, digits. = 2L) {
    format(x, digits = digits., big.mark = ",")
}
```

> Please note: This document uses count data on fatal police shootings.

```{r load-pkg, message = FALSE}
library(psych)
library(tidyverse)
library(bayesplot)
library(posterior)
theme_set(theme_classic() +
    theme(panel.grid.major.y = element_line(color = "grey92")))
```

I came across this data set from https://andrewpwheeler.com/2021/01/11/checking-a-poisson-distribution-fit-an-example-with-officer-involved-shooting-deaths-wapo-data-r-functions/

As explained [here](https://github.com/washingtonpost/data-police-shootings), the data are by the Washington Post in an effort to record every fatal shooting in the United States by a police officer since January 1, 2015.

# Research Question

> What's the rate of fatal police shootings in the United States per year?

# Data Import and Pre-Processing

```{r oid}
# Import data
fps_dat <- read_csv(
    "https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/fatal-police-shootings-data.csv"
)
```

We first count the data by year

```{r count-by-year}
# Create a year column
fps_dat <- fps_dat %>%
    mutate(year = format(date, format = "%Y"))
# Filter out 2022
fps_1521 <- filter(fps_dat, year != "2022")
count(fps_1521, year)
```

# Poisson Model

Our interest is the rate of occurrence of fatal police shootings per year. Denote this as $\theta$. The *support* of $\theta$ is $[0, \infty)$.

A Poisson model is usually a starting point for analyzing count data in a fixed amount of time. It assumes that the data follow a Poisson distribution with a fixed rate parameter:
$$P(y \mid \theta) \propto \theta^y \exp(- \theta),$$
where the data can be any non-negative integers (no decimals).

## Choosing a Prior

The [Gamma distribution](https://en.wikipedia.org/wiki/Gamma_distribution) has support: $[0, \infty)$, and is a conjugate family to the Poisson model. The Gamma distribution has the form
$$P(\theta) \propto \theta^{a - 1} \exp(-b \theta),$$
where $a$ is the prior incidence rate, and $b$ is the number of prior data points to control for the prior strength. Here, without much prior knowledge, I would simply guess there is one fatal shooting per state per month, so 600 shootings per year, but my belief is pretty weak, so I will assume a prior $b$ of 1 / 200 (one observation is one year). The $a$ will be 600 * $b$ = 3.

Here's a plot:

```{r prior-gamma}
ggplot(tibble(th = c(0, 2000)), aes(x = th)) +
    stat_function(fun = dgamma,
    args = list(shape = 3, rate = 1 / 200)) +
    labs(y = "", x = expression(theta))
```

### Prior predictive check

```{r prior-check}
set.seed(2200)
num_draws <- 100
sim_theta <- rgamma(num_draws, shape = 3, rate = 1 / 200)
# Initialize a time-series plot
ts_plot <- ggplot(count(fps_1521, year), aes(x = year, y = n))
# Simulate time series from the prior, and add to plot
# Initialize an S by N matrix to store the simulated data
y_tilde <- matrix(NA,
                  nrow = length(sim_theta),
                  ncol = 7)
colnames(y_tilde) <- 2015:2021  # add column names
for (s in seq_along(sim_theta)) {
    theta_s <- sim_theta[s]
    y_tilde[s,] <- rpois(7, lambda = theta_s)
}
# Convert to a data frame
y_tilde_df <- as_tibble(y_tilde) %>%
    rownames_to_column(var = "id") %>%
    pivot_longer(
        cols = -id,
        names_to = "year",
        values_to = "y_tilde"
    )
ts_plot +
    geom_line(
        data = y_tilde_df,
        aes(y = y_tilde, group = id),
        alpha = 0.3,
        col = "skyblue"
    ) +
    # add observed data
    geom_point() +
    geom_line(aes(group = 1))
```

## Posterior

With a conjugate prior, the posterior distribution is Gamma($a$ + $\sum_{i = 1}^N y_i$, $b$ + $N$).

```{r posterior-gamma}
ggplot(tibble(th = c(0, 2000)), aes(x = th)) +
    stat_function(fun = dgamma,
    args = list(shape = 3 + nrow(fps_1521),
                rate = 1 / 200 + 7), n = 501) +
    labs(y = "", x = expression(theta))
```

### Summary

Use simulation

```{r post-summ}
num_draws <- 1000
post_theta <- rgamma(num_draws, shape = 3 + nrow(fps_1521),
                    rate = 1 / 200 + 7)
# Use the `posterior` package
posterior::as_draws(tibble(theta = post_theta)) %>%
    summarize_draws()
```

So the estimated rate is 987 per year, with a 90% CI [967, 1006].

### Posterior Predictive Check

```{r ppc-time-series}
# Initialize a time-series plot
ts_plot <- ggplot(count(fps_1521, year), aes(x = year, y = n))
# Simulate time series from the prior, and add to plot
# Initialize an S by N matrix to store the simulated data
y_tilde <- matrix(NA,
                  nrow = length(post_theta),
                  ncol = 7)
colnames(y_tilde) <- 2015:2021  # add column names
for (s in seq_along(post_theta)) {
    theta_s <- post_theta[s]
    y_tilde[s,] <- rpois(7, lambda = theta_s)
}
# Convert to a data frame
y_tilde_df <- as_tibble(y_tilde) %>%
    rownames_to_column(var = "id") %>%
    pivot_longer(
        cols = -id,
        names_to = "year",
        values_to = "y_tilde"
    )
ts_plot +
    geom_line(
        data = y_tilde_df,
        aes(y = y_tilde, group = id),
        alpha = 0.2,
        col = "skyblue"
    ) +
    # add observed data
    geom_point() +
    geom_line(aes(group = 1))
```
