---
title: "HW 9 Template"
shorttitle: "HW 9"
author:
  - name: "Mark Lai"
    affiliation: "University of Southern California"
authornote:
abstract:
csl: "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass: "apa7"
output: papaja::apa6_pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-pkg, message = FALSE}
library(ggplot2)  # for plots
library(magrittr)  # for `%>%` operator
library(here)
library(readxl)  # for reading excel files
library(modelsummary)  # for summarizing data
library(rstan)
rstan_options(auto_write = TRUE)  # save compiled STAN object
options(mc.cores = 2)  # use two cores
library(posterior)
library(bayesplot)
theme_set(theme_classic() +
    theme(panel.grid.major.y = element_line(color = "grey92")))
```

# Research Question

> Is there difference in student evaluation rating for native and non-native English-speaking instructors?

# Variables

- `profevaluation`: evaluation rating of the instructor: 1 (very unsatisfactory) to 5 (excellent)
- `nonenglish`: 1 = non-native English speakers, 0 = native-English speakers

## Import Data

```{r beauty}
beauty <- read_excel(here("data", "ProfEvaltnsBeautyPublic.xls"))
```

## Variable Summary

```{r summ-var}
datasummary(profevaluation  * 
                (N + Mean + SD + Min + Max + Histogram) ~ 
                factor(nonenglish, labels = c("Native", "Non-Native")),
            data = beauty)
```

# Model

Let $Y$ = profevaluation, $G$ = nonenglish

Model:
$$
  \begin{aligned}
    Y_{i, G = 0} & \sim N(\mu_1, \sigma_1) \\
    Y_{i, G = 1} & \sim N(\mu_2, \sigma_2)
  \end{aligned}
$$

Prior:
$$
  \begin{aligned}
    \mu_1 & \sim N(3, 2) \\
    \mu_2 & \sim N(3, 2) \\
    \sigma_1 & \sim N^+(0, 2) \\
    \sigma_2 & \sim N^+(0, 2)
  \end{aligned}
$$

## Running Stan

We used 4 chains, each with 4,000 iterations (first 2,000 as warm-ups). 

```{r m1, results = "hide"}
# 1. form the data list for Stan
stan_dat <- with(beauty,
    list(N1 = sum(nonenglish == 0),
         N2 = sum(nonenglish == 1),
         y1 = profevaluation[which(nonenglish == 0)],
         y2 = profevaluation[which(nonenglish == 1)])
)
# 2. Run Stan
m1 <- stan(
    file = here("hw/stan", "normal_2group.stan"),
    data = stan_dat,
    seed = 1234,  # for reproducibility
    iter = 4000
)
```

# Results

As shown in the graph below, the chains mixed well.

```{r rank-hist-m1}
mcmc_rank_hist(m1, pars = c("mu1", "mu2", "sigma1", "sigma2"))
```

The following table shows the posterior distributions of $\mu_1$, $\mu_2$, $\sigma_1$, $\sigma_2$, and $\mu_2 - \mu_1$.

```{r summ-m1}
summ_m1 <- as_draws_df(m1) %>%
    subset_draws(variable = c("mu1", "mu2", "sigma1", "sigma2")) %>%
    mutate_variables(`mu2 - mu1` = mu2 - mu1) %>%
    summarise_draws()
knitr::kable(summ_m1, digits = 2)
```

The analysis showed that on average, non-native speaking instructors received a lower evaluation rating than native speaker instructors, with a posterior mean of `r round(summ_m1$mean[5], 2)` and a 90% CI of [`r round(summ_m1$q5[5], 2)`, `r round(summ_m1$q95[5], 2)`].
